Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-03-18 paper/paper.tex
Stuff

\section{Formalising Recursion Schemes}
Recursion schemes provide an abstract way to consume and generate data. We know
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise algebras for container extensions (Section~\ref{sec:algebras}), then
we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together
these notions to formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:rec-coalgebras}).

@@

Given a type \coq{A} and a functor \coq{F}, an \coq{F}-algebra is a pair given
by a type \coq{A} called the \emph{carrier} of a \emph{structure map} of
type \coq{App F A ~> A}:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
The least fixed-point for a functor \coq{F} is an instance of an \coq{F}-algebra
where the structure map is an isomorphism. This is sometimes referred to as the
\emph{initial algebra} for a functor \coq{F}. We will explain the reason behind this name shortly. \footnote[1]{To be more precise,
initial algebras are isomorphisms by Lambek's lemma, but isomorphisms of
functors do not necessarily correspond to initial algebras.}  We define the
least fixed-point of \coq{F} as an inductive type:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
\mpav{David, make it more Coq syntax like}
where \coq{LFix_in} is the \coq{F}-algebra while \coq{LFix_out} is its inverse.
As an example, the initial \coq{F}-algebra for the
functor \coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
operation \coq{Cons : A * LFix F -> LFix F}.

@@

\subsection{Coalgebras and Anamorphisms}
\label{sec:coalg}
Algebras and catamorphisms dualise nicely to the coinductive setting.  The dual
of an algebra is a coalgebra. A coalgebra can be thought of as an
\emph{observation} map. For example, for a type of states \coq{X} and an
alphabet type \coq{L} we can define a labelled transition system (LTS)
on \coq{X} as a function \coq{c : X -> FX} for a functor \coq{FX = L * X}
implementing the \emph{transition} map. In particular, for a
state \coq{x1 : X}, \coq{c x1} returns a pair \coq{(l, x2) : L * X}
where \coq{l : L} is the observable action and \coq{x2 : X} is the next state.

In general, for a functor \coq{F}, an \coq{F}-coalgebra is a pair of a
type \coq{X} called the \emph{carrier} of the coalgebra and a structure
map \coq{X -> FX}.

@@

Dually to the initial \coq{F}-algebra, a final \coq{F}-coalgebra is the greatest
fixed-point for a functor \coq{F}. We define it using a coinductive data type:
\begin{coqcode}
CoInductive GFix `(F : Cont Sh Po) : Type := GFix_in { GFix_out : App F GFix }.
\end{coqcode}
where \coq{GFix_out} is the final \coq{F}-coalgebra and \coq{GFix-in} is its
inverse witnessing the isomorphism. Similarly to \coq{LFix}, \coq{GFix} is also
defined as a setoid, with an equivalence relation that is the greatest fixpoint
of \coq{AppR}. Additionally, we define smart constructors for the isomorphism of
greatest fixed points
\begin{coqcode}
g_in : App F (GFix F) ~> GFix F                g_out : GFix F ~> App F (GFix F)
\end{coqcode}

@@

\mpav{David, I need full def of this ana}
We define anamorphisms as follows:
\begin{coqcode}
ana : forall `{F : Cont Sh P} `{setoid A}, Coalg F A ~> A ~> GFix F
\end{coqcode}
From this definition the universality property falls out:
\begin{coqcode}
Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
: f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
In words, for any \coq{F}-coalgebra, if there is any other function \coq{f} that is a \coq{F}-coalgebra homomorphism then it must be the anamorphism on the same coalgebra.

@@

Thus, in general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the
unique solution (when it exists) to the equation
\begin{equation}
  \label{eq:hylo}
  f = a \circ F\; f \circ c
\end{equation}
However this solution does not necessarily exist for an arbitrary pair of
algebras and coalgebras. In fact, this definition cannot be accepted by Coq,
since there is no structurally smaller argument to do recursion on and there is
no corecursive definition that is guaranteed to be syntactically guarded -- as
required by Coq -- due to the use of an arbitrary algebra $a$.
To address these issues, we mechanise \emph{recursive
hylomorphisms} which are guaranteed to have a unique solution to the
hylomorphism equation. These are hylomoprhisms where the coalgebra is
\emph{recursive}, i.e. coalgebras that terminate on all inputs. The following
predicate captures that a coalgebra terminates on an input: 
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
A recursive coalgebra of type \coq{RCoalg F A} is a coalgebra 
\coq{c : Coalg F A} such that \coq{forall x, RecF c x}.
Recursive hylomorphisms are implemented in Coq recursively on the structure of
the proof for the type (\coq{RecF}) as follows:
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A) 
  : forall (x : A), RecF c x -> B := fix f x H
  := match c x as h0 return (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
     with | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
     end (RecF_inv H).
\end{coqcode}
We use \coq{RecF_inv} to obtain the structurally smaller proof to use in
the recursive calls. As we did with catamorphisms and anamorphisms, we
prove that \coq{hylo_def} is respectful, and use this proof to build the
corresponding higher-order proper morphism:
\begin{coqcode}
  hylo : forall `{F : Cont Sh P} `{setoid A} `{setoid B}, 
    Alg F B ~> RCoalg F A ~> A ~> B
\end{coqcode}
Finally, we show that recursive hylomorphisms are the unique solution to the
hylomorphism equation.
\begin{coqcode}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{coqcode}
Hylomorphisms fusion falls out from the universal property:
\begin{coqcode}
Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
  (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.
\end{coqcode}
% Qed.
%   rewrite hylo_univ, fmap_comp, compA, <- E2,
%           <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.

@@

\subsubsection{On the subtype of finite elements}
In this development we have defined recursive anamorphisms on inductive data
types.  We might have as well defined them on the subtype of finite elements of
coinductive data types using a predicate which states when an element of a
coinductive data type is finite:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Now the subtype \coq{{x : GFix F | FinF x}} of finite elements for \coq{GFix F}
is isomorphic its corresponding the inductive data type \coq{LFix F}. This is
easy to see. We first define a catamorphism \coq{ccata_f_} from the
subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
any \coq{F}-algebra.
\begin{coqcode}
Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
  : forall x : GFix F, FinF x -> A := fix f x H :=
    let hx := g_out x in
      g (MkCont (shape hx) (fun e => f (cont hx e) (FinF_inv H e))).
\end{coqcode}
We now prove this is isomorphic to the least fixed-point of the functor \coq{F}
We take the catamorphism from the finite elements of \coq{GFix F} to the
inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}. Its inverse is the catamorphism on the restriction of \coq{g_in} to the finite elements of \coq{GFix}, which we denote by \coq{lg_in}.
The following lemmas prove the isomorphism:
\begin{coqcode}
Lemma cata_ccata `{setoid A} : cata lg_in \o ccata l_in =e id.
Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata l_in \o cata lg_in =e id.
\end{coqcode}
% This predicate is equivalent to the proposition that
% the anamorphism of such coalgebras produce only finite trees:
% \begin{coqcode}
% Lemma ana_rec_term `{setoid A} (h : Coalg F A)
%   : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
% \end{coqcode}

@@

Both mergesort and quicksort are divide-and-conquer algorithms that can be
captured by the structure of an hylomorphisms. The structure of the recursion
is that of a binary tree.  For example, in the case of quicksort, a list is
split into a pivot, the label of the node, and two sublists. We define the data
functor of trees as follows:
\begin{coqcode}
Inductive ITreeF L N X := i_leaf (l : L) | i_node (n : N) (l r : X)
\end{coqcode}
We define the functor as a container using the following shapes and positions:
\begin{coqcode}
Inductive Tshape L A := | Leaf (ELEM : L) | Node (ELEM : A).
Inductive Tpos := | Lbranch | Rbranch. 
\end{coqcode}
These define a container, \coq{TreeF}, in a straightforward way, by making
the positions of type \coq{Tpos} only valid in \coq{Node}.
We define a series of definitions for tree container constructors and 
destructors:
\mpav{The following algebra should be called a\_in or l\_in whatever}
\begin{coqcode}
Definition a_out {L A X : Type} : App (TreeF L A) X ~> ITreeF L A X.

(* Converting TreeF to ITreeF *)
Notation a_leaf x := (MkCont (Leaf _ x) (@dom_leaf _ _ _ x)).
Notation a_node x l r := (MkCont (Node _ x)
  (fun p => match val p with | Lbranch => l | Rbranch => r end)).
\end{coqcode}

% \subsubsection{Quicksort}

The container for the Quicksort hylomorphism is \coq{TreeF unit int}, with
the following algebra and coalgebra.
\begin{coqcode}
  Definition merge : App (TreeF unit int) (list int) ~> list int.
|{ x : (App (TreeF unit int) (list int)) ~> (
           match x with
           | MkCont sx kx =>
               match sx return (Container.Pos sx -> _) -> _ with
               | Leaf _ _ => fun _ => nil
               | Node _ h => fun k => List.app (k (posL h)) (h :: k (posR h))
               end kx
           end
  )}|.
Defined.

@@

Definition Table := LFix Memo.
\end{coqcode}
Memoisation tables are the least fixed point of the container defined by shapes
\coq{A * Sg} and positions \coq{Pg}, given a a container \coq{G : Cont Sg Pg}.
We define a function to insert elements into the memoisation tables:
\begin{coqcode}
  Definition Cons : A * App G Table ~> App Memo Table := (* *)
\end{coqcode}
And two functions to inspect the head of a memoisation table, and remove an
element from it:
\begin{coqcode}
Definition headT : Table ~> A := (* *)
Definition tailT : Table ~> App G Table := (* *)
\end{coqcode}
These tables map ``paths'' in the least fixed point of \coq{Memo} to elements
of type \coq{A}. For example, if \coq{G} is a list-generating functor, these
paths will be natural numbers. 
Using these definitions, a dynamorphism is defined as follows:
\begin{coqcode}
Definition dyna (a : App G Table ~> A) (c : RCoalg G B) : B ~> A
:= headT \o hylo (l_in \o Cons \o pair a id) c.
\end{coqcode}
Note how, instead of an algebra \coq{App G A ~> A}, the algebra takes a
memoisation table. The definition of the algebra can use this table to lookup
elements, instead of triggering a further recursive call. Elements are inserted
into the memoisation table by the use of \coq{Cons} to the result of applying
the algebra. The algebra for the dynamorphism looks up the previously computed
elements to produce the result, thus saving the corresponding recursive calls:
\begin{coqcode}
Fixpoint memo_knap table wvs :=
  match wvs with | nil => nil | h :: t =>
      match lookupT (Nat.pred (fst h)) table with
      | Some u => (u + snd h)%sint63 :: memo_knapsack table t
      | None => memo_knapsack table t
      end
  end.

@@

\subsection{Shortcut Deforestation}\label{sec:shortcut}
The final case study we consider is shortcut deforestation on lists.  Shortcut
deforestation can be expressed succintly in terms of hylomorphisms and their
laws~\cite{TakanoM95}. In particular, given a function:
\begin{coqcode}
  s  : forall A. (App F A -> A) -> (App F A -> A)
\end{coqcode}
We can conclude, by parametricity, that
\begin{coqcode}
  hylo a l_out \o hylo (sigma l_in) c =e hylo (s a) c
\end{coqcode}
This is generally known as the \emph{acid rain} theorem. Unfortunately, this is
\textbf{not} provable in Coq if we want to remain axiom-free, since we would
need to add the necessary parametricity
axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.

However, we prove a specific version of this theorem for the list generating
container (i.e.\ the container whose least fixed point is a list), and use it
to encode the example from Takano and Meijer~\cite{TakanoM95}:
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
\end{coqcode}
Here, we are defining function \coq{sf1} as the composition of \coq{length},
\coq{Lmap f} and \coq{append ys}. Functions \coq{length} and \coq{Lmap} are
catamorphisms. Function \coq{append ys} is also a catamorphism that
appends \coq{ys} to an input list. It is defined by applying an algebra to
every cons node of a list, and applying a catamorphism with the input algebra
to \coq{ys} in the nil case:
\begin{coqcode}
Definition tau (l : list A) (a : Alg (ListF A) B) : App (ListF A) B -> B :=
  fun x => match x with | MkCont sx kx => match sx with
  | s_nil => fun _ => (hylo a ilist_coalg) l
  | s_cons h => fun kx => a (MkCont (s_cons h) kx)
  end kx end.
Definition append (l : list A) := hylo (tau l l_in) ilist_coalg.
\end{coqcode}
Here, \coq{ilist_coalg} is a recursive coalgebra from Coq lists to the
\coq{ListF} container. We apply the hylo fusion law repeatedly, unfold
definitions, and simplify in our specification for \coq{sf1}:
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
  rewrite hylo_map_fusion, <- acid_rain. simpl; reflexivity.
Defined.
\end{coqcode}
From this, we extract the following OCaml code:
\begin{ocamlcode}
let rec sf1 f ys = function | [] -> let rec f0 = function
                                      | [] -> (Uint63.of_int (0))
                                      | _ :: t -> add (Uint63.of_int (1)) (f0 t)
                                    in f0 ys
                            | _ :: t -> add (Uint63.of_int (1)) (sf1 f ys t)
\end{ocamlcode}

@@

\section{On Equality}
This definition highlights why we cannot define equality of the positions and
the contained elements generically (recall Section~\ref{sec:TODO}): the
contained types may not have (yet) an associated equivalence. In other words, we
would need to provide an equivalence for \coq{LFix}, but the only way to do this
is to define such equivalence simultaneously to the definition for  \coq{LFix}.
Such definition would be rejected by Coq. There are other approaches that can
circumvent this problem, by separating the definition of least fixed points from
the proofs that the functions that get the contained elements are proper
morphisms. But this leads to longer, unnecessarily complicated proofs.
Furthermore, this extra layer interferes with code extraction, in that it
requires destructing/constructing subsequent \coq{LFix} and \coq{App}, and Coq
may not always able to inline them.

\mpav{Dumped here}
We define the equality of initial algebras of container extensions recursively,
by traversing the structure and equating the shapes and positions. This
definition is equivalent to taking the fixpoint of \coq{AppR} in
Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
positions, the contents must be related by \coq{LFixR}.
\begin{coqcode}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{coqcode}

>>>>

\section{Formalising Recursion Schemes}
Recursion schemes provide an abstract way to consume and generate data. We now
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise algebras for container extensions (Section~\ref{sec:algebras}), then
we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together
these notions to formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:rec-coalgebras}).

@@

Given a type \coq{A} and a functor \coq{F}, an \coq{F}-algebra is a pair given
by a type \coq{A} called the \emph{carrier} of a \emph{structure map} of
type \coq{App F A ~> A}:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
The least fixed-point for a functor \coq{F} is an instance of an \coq{F}-algebra
where the structure map is an isomorphism. This is sometimes referred to as the
\emph{initial algebra} for a functor \coq{F}. We will explain the reason behind this name shortly. \footnote[1]{To be more precise,
initial algebras are isomorphisms by Lambek's lemma, but isomorphisms of
functors do not necessarily correspond to initial algebras.}  We define the
least fixed-point of \coq{F} as an inductive type:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
where \coq{LFix_in} is the \coq{F}-algebra while \coq{LFix_out} is its inverse.
As an example, the initial \coq{F}-algebra for the
functor \coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
operation \coq{Cons : A * LFix F -> LFix F}.

@@

\subsection{Coalgebras and Anamorphisms}
\label{sec:coalg}
Algebras and catamorphisms dualise nicely to the coinductive setting.  The dual
of an algebra is a coalgebra. A coalgebra can be thought of as an
\emph{observation} map. For example, for a type of states \coq{X} and an
alphabet type \coq{L} we can define a labelled transition system (LTS)
on \coq{X} as a function \coq{c : X -> FX} for a functor \coq{F X = L * X}
implementing the \emph{transition} map. In particular, for a
state \coq{x1 : X}, \coq{c x1} returns a pair \coq{(l, x2) : L * X}
where \coq{l : L} is the observable action and \coq{x2 : X} is the next state.

In general, for a functor \coq{F}, an \coq{F}-coalgebra is a pair of a
type \coq{X} called the \emph{carrier} of the coalgebra and a structure
map \coq{X -> F X}.

@@

Dually to the initial \coq{F}-algebra, a final \coq{F}-coalgebra is the greatest
fixed-point for a functor \coq{F}. We define it using a coinductive data type:
\begin{coqcode}
CoInductive GFix `(F : Cont Sh Po) : Type := GFix_in { GFix_out : App F GFix }.
\end{coqcode}
where \coq{GFix_out} is the final \coq{F}-coalgebra and \coq{GFix_in} is its
inverse witnessing the isomorphism. Similarly to \coq{LFix}, \coq{GFix} is also
defined as a setoid, with an equivalence relation that is the greatest fixpoint
of \coq{AppR}. Additionally, we define smart constructors for the isomorphism of
greatest fixed points
\begin{coqcode}
g_in : App F (GFix F) ~> GFix F                g_out : GFix F ~> App F (GFix F)
\end{coqcode}

@@

We define anamorphisms as follows:
\begin{coqcode}
Definition ana_f_ (c : Coalg F A) :=
  cofix f x := 
    match c x with | MkApp sx kx => GFix_in (MkApp sx (fun e => f (kx e))) end.

Definition ana : forall `{setoid A}, Coalg F A ~> A ~> GFix F := (*... ana_f_ ... *)
\end{coqcode}
From this definition the universality property falls out:
\begin{coqcode}
Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix F)
: f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
In words, for any \coq{F}-coalgebra, if there is any other function \coq{f}
that is a \coq{F}-coalgebra homomorphism then it must be the anamorphism on the
same coalgebra.

@@

Thus, in general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the
unique solution (when it exists) to the equation
\begin{equation}
  \label{eq:hylo}
  f = a \circ F\; f \circ c
\end{equation}
However this solution does not necessarily exist for an arbitrary pair of
algebras and coalgebras. In fact, this definition cannot be accepted by Coq,
since there is no structurally smaller argument to do recursion on and there is
no corecursive definition that is guaranteed to be syntactically guarded -- as
required by Coq -- due to the use of an arbitrary algebra $a$.
To address these issues, we mechanise \emph{recursive
hylomorphisms} which are guaranteed to have a unique solution to the
hylomorphism equation. These are hylomoprhisms where the coalgebra is
\emph{recursive}, i.e. coalgebras that terminate on all inputs. The following
predicate captures that a coalgebra terminates on an input: 
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
A recursive coalgebra of type \coq{RCoalg F A} is a coalgebra 
\coq{c} such that \coq{forall x, RecF c x}.
Recursive hylomorphisms are implemented in Coq recursively on the structure of
the proof for the type (\coq{RecF}) as follows:
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A) 
  : forall (x : A), RecF c x -> B := fix f x H
  := match c x as h0 return (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
     with | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
     end (RecF_inv H).
\end{coqcode}
We use \coq{RecF_inv} to obtain the structurally smaller proof to use in
the recursive calls. As we did with catamorphisms and anamorphisms, we
prove that \coq{hylo_def} is respectful, and use this proof to build the
corresponding higher-order proper morphism:
\begin{coqcode}
  hylo : forall `{F : Cont Sh P} `{setoid A} `{setoid B}, 
    Alg F B ~> RCoalg F A ~> A ~> B
\end{coqcode}
Finally, we show that recursive hylomorphisms are the unique solution to the
hylomorphism equation.
\begin{coqcode}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{coqcode}
Hylomorphisms fusion falls out from the universal property:
\begin{coqcode}
Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
  (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.
\end{coqcode}
% Qed.
%   rewrite hylo_univ, fmap_comp, compA, <- E2,
%           <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.

@@

\subsubsection{On the subtype of finite elements}
In this development we have defined recursive anamorphisms on inductive data
types.  We might have as well defined them on the subtype of finite elements of
coinductive data types using a predicate which states when an element of a
coinductive data type is finite:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Now the subtype \coq{{x : GFix F | FinF x}} of finite elements for \coq{GFix F}
is isomorphic its corresponding the inductive data type \coq{LFix F}. This is
easy to see. We first define a catamorphism \coq{ccata_f_} from the
subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
any \coq{F}-algebra.
\begin{coqcode}
Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
  : forall x : GFix F, FinF x -> A := fix f x H :=
    let hx := g_out x in
      g (MkCont (shape hx) (fun e => f (cont hx e) (FinF_inv H e))).
\end{coqcode}
We now prove this is isomorphic to the least fixed-point of the functor
\coq{F}.
We take the catamorphism from the finite elements of \coq{GFix F} to the
inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}. Its
inverse is the catamorphism on the restriction of \coq{g_in} to the finite
elements of \coq{GFix}, which we denote by \coq{lg_in}.  The following lemmas
prove the isomorphism:
\begin{coqcode}
Lemma cata_ccata `{setoid A} : cata lg_in \o ccata l_in =e id.
Lemma ccata_cata `{setoid A} : ccata l_in \o cata lg_in =e id.
\end{coqcode}
The finite subtype of \coq{GFix F} allows us to compose
catamorphisms and anamorphisms, by using the above isomorphism. In our work,
however, we use \emph{recursive} anamorphisms, defined as \coq{hylo l_in c} for
a recursive coalgebra \coq{c}, which compose easily with
catamorphisms.
% This predicate is equivalent to the proposition that
% the anamorphism of such coalgebras produce only finite trees:
% \begin{coqcode}
% Lemma ana_rec_term `{setoid A} (h : Coalg F A)
%   : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
% \end{coqcode}

@@

Both mergesort and quicksort are divide-and-conquer algorithms that can be
captured by the structure of an hylomorphisms. The structure of the recursion
is that of a binary tree.  For example, in the case of quicksort, a list is
split into a pivot, the label of the node, and two sublists. We define the data
functor of trees as follows:
\begin{coqcode}
Inductive ITreeF L N X := i_leaf (l : L) | i_node (n : N) (l r : X)
\end{coqcode}
We define the functor as a container using the following shapes and positions:
\begin{coqcode}
Inductive Tshape L A := | Leaf (ELEM : L) | Node (ELEM : A).
Inductive Tpos := | Lbranch | Rbranch. 
\end{coqcode}
These define a container, \coq{TreeF}, in a straightforward way, by making
the positions of type \coq{Tpos} only valid in \coq{Node}.
We define a series of definitions for tree container constructors and 
destructors:
% \mpav{The following algebra should be called a\_in or l\_in whatever}
\begin{coqcode}
Definition a_out {L A X : Type} : App (TreeF L A) X ~> ITreeF L A X.
Notation a_leaf x := (MkCont (Leaf _ x) (@dom_leaf _ _ _ x)).
Notation a_node x l r := (MkCont (Node _ x)
  (fun p => match val p with | Lbranch => l | Rbranch => r end)).
\end{coqcode}
The container for the Quicksort hylomorphism is \coq{TreeF unit int}, with
the following algebra and coalgebra.
\begin{coqcode}
  Definition merge : App (TreeF unit int) (list int) ~> list int.
|{ x : (App (TreeF unit int) (list int)) ~> (
           match x with
           | MkCont sx kx =>
               match sx return (Container.Pos sx -> _) -> _ with
               | Leaf _ _ => fun _ => nil
               | Node _ h => fun k => List.app (k (posL h)) (h :: k (posR h))
               end kx
           end
  )}|.
Defined.

@@

Definition Table := LFix Memo.
\end{coqcode}
Memoisation tables are the least fixed point of the container defined by shapes
\coq{A * Sg} and positions \coq{Pg}, given a a container \coq{G : Cont Sg Pg}.
We define a function to insert elements into the memo tables:
\begin{coqcode}
  Definition Cons : A * App G Table ~> App Memo Table := (* *)
\end{coqcode}
And two functions to inspect the head of a memo table, and remove an
element from it:
\begin{coqcode}
Definition headT : Table ~> A := (* *)
Definition tailT : Table ~> App G Table := (* *)
\end{coqcode}
These tables map ``paths'' in the least fixed point of \coq{Memo} to elements
of type \coq{A}. For example, if \coq{G} is a list-generating functor, these
paths will be natural numbers. 
Using these definitions, a dynamorphism is defined as follows:
\begin{coqcode}
Definition dyna (a : App G Table ~> A) (c : RCoalg G B) : B ~> A
:= headT \o hylo (l_in \o Cons \o pair a id) c.
\end{coqcode}
Note how, instead of an algebra \coq{App G A ~> A}, the algebra takes a
memo table. The definition of the algebra can use this table to lookup
elements, instead of triggering a further recursive call. Elements are inserted
into the memoisation table by the use of \coq{Cons} to the result of applying
the algebra. The algebra for the dynamorphism looks up the previously computed
elements to produce the result, thus saving the corresponding recursive calls:
\begin{coqcode}
Fixpoint memo_knap table wvs :=
  match wvs with | nil => nil | h :: t =>
      match lookupT (Nat.pred (fst h)) table with
      | Some u => (u + snd h)%sint63 :: memo_knapsack table t
      | None => memo_knapsack table t
      end
  end.

@@

\subsection{Shortcut Deforestation}\label{sec:shortcut}
The final case study we consider is shortcut deforestation on lists.  Shortcut
deforestation can be expressed succintly in terms of hylomorphisms and their
laws~\cite{TakanoM95}. In particular, given a function:
\begin{coqcode}
  s  : forall A. (App F A -> A) -> (App F A -> A)
\end{coqcode}
We can conclude, by parametricity, that
\begin{coqcode}
  hylo a l_out \o hylo (sigma l_in) c =e hylo (s a) c
\end{coqcode}
This is generally known as the \emph{acid rain} theorem. Unfortunately, this is
\textbf{not} provable in Coq if we want to remain axiom-free, since we would
need to add the necessary parametricity
axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.
However, we prove a specific version of this theorem for the list generating
container (i.e.\ the container whose least fixed point is a list), and use it
to encode the example from Takano and Meijer~\cite{TakanoM95}:
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
\end{coqcode}
Here, we are defining function \coq{sf1} as the composition of \coq{length},
\coq{Lmap f} and \coq{append ys}. Functions \coq{length} and \coq{Lmap} are
catamorphisms. Function \coq{append ys} is also a catamorphism that
appends \coq{ys} to an input list. It is defined by applying an algebra to
every cons node of a list, and applying a catamorphism with the input algebra
to \coq{ys} in the nil case:
\begin{coqcode}
Definition tau (l : list A) (a : Alg (ListF A) B) : App (ListF A) B -> B :=
  fun x => match x with | MkCont sx kx => match sx with
  | s_nil => fun _ => (hylo a ilist_coalg) l
  | s_cons h => fun kx => a (MkCont (s_cons h) kx)
  end kx end.
Definition append (l : list A) := hylo (tau l l_in) ilist_coalg.
\end{coqcode}
Here, \coq{ilist_coalg} is a recursive coalgebra from Coq lists to the
\coq{ListF} container. We apply the hylo fusion law repeatedly, unfold
definitions, and simplify in our specification for \coq{sf1}:
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
  rewrite hylo_map_fusion, <- acid_rain. simpl; reflexivity.
Defined.
\end{coqcode}
From this, we extract the following OCaml code:
\begin{ocamlcode}
let rec sf1 f ys = function | [] -> let rec f0 = function
                                      | [] -> (Uint63.of_int (0))
                                      | _ :: t -> add (Uint63.of_int (1)) (f0 t)
                                    in f0 ys
                            | _ :: t -> add (Uint63.of_int (1)) (sf1 f ys t)
\end{ocamlcode}

@@

\section{On Equality}
This definition highlights why we cannot define equality of the positions and
the contained elements generically: the
contained types may not have (yet) an associated equivalence. In other words, we
would need to provide an equivalence for \coq{LFix}, but the only way to do this
is to define such equivalence simultaneously to the definition for  \coq{LFix}.
Such definition would be rejected by Coq. There are other approaches that can
circumvent this problem, by separating the definition of least fixed points from
the proofs that the functions that get the contained elements are proper
morphisms. But this leads to longer, unnecessarily complicated proofs.
Furthermore, this extra layer interferes with code extraction, in that it
requires destructing/constructing subsequent \coq{LFix} and \coq{App}, and Coq
may not always able to inline them.

\mpav{Dumped here}
We define the equality of initial algebras of container extensions recursively,
by traversing the structure and equating the shapes and positions. This
definition is equivalent to taking the fixpoint of \coq{AppR} where the shapes
must be equivalent and, for equal positions, the contents must be related by
\coq{LFixR}.
\begin{coqcode}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{coqcode}
